\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-11-15 k-planar.tex
nada
minor changes implemented

\begin{abstract}
Dujmovi\'c~et~al.~[\emph{J.~ACM}~'20] recently proved that every planar graph is isomorphic to a subgraph of the strong product of a bounded treewidth graph and a path. Analogous results were obtained for graphs of bounded Euler genus or apex-minor-free graphs. These tools have been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and adjacency labelling. This paper proves analogous product structure theorems for various non-minor-closed classes. One noteable example is $k$-planar graphs (those with a drawing in the plane in which each edge is involved in at most $k$ crossings). We prove that every $k$-planar graph is isomorphic to a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a more general setting based on so-called shortcut systems, which are of independent interest. This leads to analogous results for certain types of map graphs, string graphs, graph powers, and nearest neighbour graphs.
\end{abstract}
\end{titlepage}
\pagenumbering{roman}
\tableofcontents
\newpage

@@

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20} (with an improvement by \citet{UWY22}). A graph $G$ is \defin{contained} in a graph $X$ if $G$ is isomorphic to a subgraph of $X$.

\begin{thm}[\citep{DJMMUW20,UWY22}]
\label{PlanarProduct}
Every planar graph is contained in:
\begin{compactenum}[(a)]
  \item $H\boxtimes P$ for some graph $H$ of treewidth at most $6$ and for some path $P$,
  \item $H\boxtimes P \boxtimes K_3$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
\end{compactenum}
\end{thm}

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to prove that planar graphs have $p$-centered chromatic number $O(p^2\log p)$ and give a matching lower bound.
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a `universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $\ell$-vertex ranking~\citep{BDJM} and twin-width~\citep{BKW}.

All of these results hold for any graph class that has a product structure theorem analogous to \cref{PlanarProduct}; that is, for any graph class  $\mathcal{G}$ where every graph in $\mathcal{G}$ is contained in $H\boxtimes P\boxtimes K_\ell$ where $H$ has bounded treewidth, $P$ is a path, and $\ell$ is bounded.\footnote{It is easily seen that $\tw(H\boxtimes K_\ell) \leq (\tw(H)+1)\ell-1$, so we may assume that $\ell=1$ in this statement.} These applications motivate finding product structure theorems for other graph classes. \citet{DJMMUW20} prove product structure theorems for graphs of bounded Euler genus\footnote{The \textit{Euler genus} of a surface with $h$ handles and $c$ crosscaps is $2h+c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} and for apex-minor-free graphs,\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \defin{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \defin{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \defin{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.} and \citet{DEMWW22} do so for bounded-degree graphs in any minor-closed class. See \citep{DHJLW21} for a survey on this topic.

@@

\begin{table}[h]
  \centering{
    \begin{tabular}{lccl}
      \hline
      {Graph class} & $\tw(H)$ & $\ell$ & {Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      planar & $4^*$ & $2$ & \cite{bose.morin.ea:optimal} \\
      planar & $6^*$ & $1$ & \cite{UWY22} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW22} \\
      genus $g$ & $2g+6$ & 1 & \cite{UWY22} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-nearest-neighbour & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \multicolumn{4}{l}{$^*$ these bounds also apply to the simple treewidth \cite{knauer.ueckerdt:simple} of $H$.}
    \end{tabular}
  }
  \caption{Product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for bounded-degree string graphs, powers of bounded-degree graphs, map graphs of bounded clique number, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure theorems and the consequences of the current work.
\end{compactitem}

@@

\begin{lem}[\citep{DJMMUW20}]
\label{PartitionProduct}
For every graph $H$, a graph $G$ has an $H$-partition of layered width at most $\ell$ if and only if $G$ is contained in $H \boxtimes P \boxtimes K_\ell$ for some path $P$.
\end{lem}

As an example of the use of layered partitions, to prove \cref{PlanarProduct}(a),
\citet{DJMMUW20} build on an earlier result of \citet{PS21} to show that every planar graph has an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$ (improved to $6$ in \citep{UWY22}). The proof is constructive and gives a simple quadratic-time algorithm for finding the corresponding partition and layering.\footnote{\citet{bose.morin.ea:optimal} have recently given linear time algorithms for computing layered $H$-partitions of planar graphs.}

@@

\noindent\textbf{Open Problem:}
  Does there exist a function $\ell:\N\to\N$ and a universal constant $C$ such that every $k$-planar graph is contained in $H\boxtimes P \boxtimes K_{\ell(k)}$ for some graph $H$ with treewidth at most $C$?  Perhaps $C=3$, which is the case for planar graphs (\cref{PlanarProduct}(b)) and for $1$-planar graphs (\cref{1_planar_product}). Note that $C\geq 3$ even for planar graphs \citep{DJMMUW20}.

@@

\begin{thm}[\citep{DJMMUW20,UWY22,DHHW22}]\label{GenusProduct}
  Every graph of Euler genus $g$ is contained in:
  \begin{compactenum}[(a)]
  \item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
  \item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some planar graph $H$ of treewidth at most $3$ and for some path $P$.
  \end{compactenum}
\end{thm}

@@

\begin{thm}
\label{gkPlanarProduct}
Every $(g,k)$-planar graph is contained in $H\boxtimes P \boxtimes K_\ell$ for some graph $H$ with $\tw(H) \leq \binom{k+4}{3}-1$, where $\ell:=\max\{2g,3\}\cdot(6k^2+16k+10)$.
% \note{DW}{I have changed $\tw(H) \leq \binom{k+5}{4}-1$ to $\tw(H) \leq \binom{k+4}{3}-1$, okay?}
%\note{PM}{Ok}
\end{thm}

@@

\note{PM}{To make the referee happy I've added the ``bounded-degree'' qualifier to string graphs and the ``bounded clique number'' qualifier to map graphs.  Do we want to explcitly state that the clique number of a $(g,d)$-map graph is bounded by a function of $g$ and $d$?  Proof (for the $g=0$ case): A $d$-map representation gives a drawing with at most $\floor{\tfrac{d-2}{2}}\ceil{\tfrac{d-2}{2}}\in O(d^2)$ crossings per edge, but any drawing of $K_{\omega}$ on a surface of genus-$g$ has an edge that is crossed $\Omega(\omega^2)$ times. So if $G$ is a $d$-map graph then $\omega(G) \in O(d)$.}

@@

For any integer $d\ge 3$, an embedded (multi)graph $G$ is \defin{$d$-framed} if it has a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in some face (possibly the outer face) of $G_0$. The embedded (multi)graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.  A $d$-framed graph $G$ is \defin{edge-maximal} with respect to its frame $G_0$ if, for each face $F$ of $G_0$ and each pair of distinct vertices $v,w\in V(F)$, $vw\in E(F)$ or $G$ contains an edge $vw$ whose interior is contained in $F$.

\note{DW}{The word ``interior'' may be misinterpreted when refering to the outerface (see referee's comment). Shall we just write ``contained in some face''? Do we need to change this elsewhere?}
\note{PM}{Done.  I've changed occurrences elsewhere also.  Under our definitions (in Section 2.3) the faces of $G$ are open sets so ``in $F$'' already means ``in the interior of $F$.''}

@@

A \defin{BFS spanning forest} $T$ of $G$ rooted at a set $V_0\subseteq V(G)$ is a spanning forest of $G$ that contains a tree rooted at each vertex of $V_0$ with the property that for each $v\in V(G)$,
\[  \min\{\dist_G(v,w):w\in V_0\}=\min\{\dist_T(v,w):w\in V_0\}.  \]
\cref{d_framed_product_stucture} is a consequence of the following technical lemma, which is an extension of the analogous result for plane graphs~\cite[Lemma~17]{DJMMUW20}.  For readers already familiar with the proof in \cite{dujmovic.joret.ea:planar}, the main difference here is that a typical tripod in the proof of \cite[Lemma~17]{dujmovic.joret.ea:planar} consists of three upward paths in a BFS tree/forest whose lower endpoints belong to a single triangular face $\tau$.  In the current setting, a typical tripod consists of three upward paths whose lower endpoints belong to a common face $\tau$ of $G_0$ plus all of the (at most $d$) vertices of $\tau$ that are not already included in tripods.

@@

\begin{proof}
This proof is similar to the proofs of Lemmas~13 and 17 in \cite{DJMMUW20}, but is complicated by several factors.  In particular, the reader should keep in mind that $G$ and $G_0$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

@@

  \begin{figure}[!b]
    \begin{center}
      \begin{tabular}{c@{}c@{}c}
        \includegraphics{figs/zoomba-1} &
        \includegraphics{figs/zoomba-2} &
        \includegraphics{figs/zoomba-3} \\
        \multicolumn{3}{c}{{\color{brew8}\raisebox{-3pt}{\rule{12pt}{12pt}}} $S$\qquad {\color{brew2}\raisebox{-3pt}{\rule{12pt}{12pt}}} $X$} \\
        $k=1$ & $k=2$ & $k=3$
      \end{tabular}
    \end{center}
    \caption{Three cases when choosing the face $\tau$ of $N_0$.}
    \label{boring_figure}
  \end{figure}

@@

\begin{proof}
Let $G$ be a (connected) $1$-plane multigraph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}. We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If distinct vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no newly-introduced edge $vw$ is on the boundary of a face with an edge parallel to $vw$.  Since $G$ is connected, each edge added during this process splits a face $F$ with a facial walk of length at least $4$ into two faces each having a shorter facial walk than that of $F$.
Therefore this process eventually reaches a stage where each face is either bounded by a $3$-cycle or a (pre-existing) $2$-cycle and the process ends.

% \note{DW}{I think $G_0$ should be replaced by $G'$ throughout this proof.}
% \note{PM}{Agreed. Done.}

@@

For an integer $k\geq 0$, a \defin{$k$-queue layout} of a graph $G$ consists of a linear ordering $\preceq$ of $V(G)$ and a partition $\{E_1,E_2,\dots,E_k\}$ of $E(G)$, such that for $i\in\{1,2,\dots,k\}$, no two edges in $E_i$ are nested with respect to $\preceq$. That is, it is not the case that $v\prec x \prec y \prec w$ for edges $vw,xy\in E_i$. The \defin{queue-number} of a graph $G$, denoted by \defin{$\qn(G)$}, is the minimum integer $k$ such that $G$ has a $k$-queue layout. Queue-number was introduced by \citet{HLR92}, who famously conjectured that planar graphs have bounded queue-number. \citet{DJMMUW20} recently proved this conjecture using \cref{PlanarProduct} and the following lemma. Indeed, resolving this question was the motivation for the development of \cref{PlanarProduct}.

\begin{lem}[\citep{DJMMUW20}]
\label{qn}
If $G\subseteq H \boxtimes P \boxtimes K_\ell$ then
$\qn(G) \leq  3 \ell \, \qn(H) + \floor{\tfrac{3}{2}\ell}
\leq 3 \ell \, 2^{\tw(H)}  - \ceil{\tfrac{3}{2}\ell}$.
\end{lem}

@@

Note that \citet{DJMMUW20} previously proved the bound of
$O(g^{k+2})$ for $(g,k)$-planar graphs using \cref{GenusProduct} and an ad-hoc method. Our result provides a better bound when $g>2^{k^2}$.

@@

For planar graphs of treewidth at most $3$ a $O(p^2\log p)$ upper bound is known:
% \note{DW}{I prefer ``a $O(p^2\log p)$'', pronounced ``a big-O...''}

>>>>

\begin{abstract}
Dujmovi\'c~et~al.~[\emph{J.~ACM}~'20] proved that every planar graph is isomorphic to a subgraph of the strong product of a bounded treewidth graph and a path. Analogous results were obtained for graphs of bounded Euler genus or apex-minor-free graphs. These tools have been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and adjacency labelling. This paper proves analogous product structure theorems for various non-minor-closed classes. One noteable example is $k$-planar graphs (those with a drawing in the plane in which each edge is involved in at most $k$ crossings). We prove that every $k$-planar graph is isomorphic to a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a more general setting based on so-called shortcut systems, which are of independent interest. This leads to analogous results for certain types of map graphs, string graphs, graph powers, and nearest neighbour graphs.
\end{abstract}
\end{titlepage}
\pagenumbering{roman}
\tableofcontents
\newpage

@@

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. Undefined terms are in \citep{Diestel5}.} by \citet{dujmovic.joret.ea:planar} (with an improvement by \citet{UWY22}). A graph $G$ is \defin{contained} in a graph $X$ if $G$ is isomorphic to a subgraph of $X$.

\begin{thm}[\citep{dujmovic.joret.ea:planar,UWY22}]
\label{PlanarProduct}
Every planar graph is contained in:
\begin{compactenum}[(a)]
  \item $H\boxtimes P$ for some graph $H$ of treewidth at most $6$ and for some path $P$,
  \item $H\boxtimes P \boxtimes K_3$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
\end{compactenum}
\end{thm}

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{dujmovic.joret.ea:planar} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to prove that planar graphs have $p$-centered chromatic number $O(p^2\log p)$ and give a matching lower bound.
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a `universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $\ell$-vertex ranking~\citep{BDJM} and twin-width~\citep{BKW}.

All of these results hold for any graph class that has a product structure theorem analogous to \cref{PlanarProduct}; that is, for any graph class  $\mathcal{G}$ where every graph in $\mathcal{G}$ is contained in $H\boxtimes P\boxtimes K_\ell$ where $H$ has bounded treewidth, $P$ is a path, and $\ell$ is bounded.\footnote{It is easily seen that $\tw(H\boxtimes K_\ell) \leq (\tw(H)+1)\ell-1$, so we may assume that $\ell=1$ in this statement.} These applications motivate finding product structure theorems for other graph classes. \citet{dujmovic.joret.ea:planar} prove product structure theorems for graphs of bounded Euler genus\footnote{The \textit{Euler genus} of a surface with $h$ handles and $c$ crosscaps is $2h+c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} and for apex-minor-free graphs,\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \defin{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \defin{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \defin{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.} and \citet{DEMWW22} do so for bounded-degree graphs in any minor-closed class. See \citep{DHJLW21} for a survey on this topic.

@@

\begin{table}[h]
  \centering{
    \begin{tabular}{lccl}
      \hline
      {Graph class} & $\tw(H)$ & $\ell$ & {Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{dujmovic.joret.ea:planar} \\
      planar & $4^*$ & $2$ & \cite{bose.morin.ea:optimal} \\
      planar & $6^*$ & $1$ & \cite{UWY22} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW22} \\
      genus $g$ & $2g+6$ & 1 & \cite{UWY22} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{dujmovic.joret.ea:planar} \\
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{dujmovic.joret.ea:planar} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-nearest-neighbour & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \multicolumn{4}{l}{$^*$ these bounds also apply to the simple treewidth \cite{knauer.ueckerdt:simple} of $H$.}
    \end{tabular}
  }
  \caption{Product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for bounded-degree 
  %\note{DW}{We actually make a stronger assumption than bounded degree; we assume each string is crossed at most $k$ times. I am happy to leave our write-up as is, but be aware of the difference. } 
  string graphs, powers of bounded-degree graphs, map graphs of bounded clique number, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure theorems and the consequences of the current work.
\end{compactitem}

@@

\begin{lem}[\citep{dujmovic.joret.ea:planar}]
\label{PartitionProduct}
For every graph $H$, a graph $G$ has an $H$-partition of layered width at most $\ell$ if and only if $G$ is contained in $H \boxtimes P \boxtimes K_\ell$ for some path $P$.
\end{lem}

As an example of the use of layered partitions, to prove \cref{PlanarProduct}(a),
\citet{dujmovic.joret.ea:planar} build on an earlier result of \citet{PS21} to show that every planar graph has an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$ (improved to $6$ in \citep{UWY22}). The proof is constructive and gives a simple quadratic-time algorithm for finding the corresponding partition and layering.\footnote{\citet{bose.morin.ea:optimal} have recently given linear time algorithms for computing layered $H$-partitions of planar graphs.}

@@

\noindent\textbf{Open Problem:}
  Does there exist a function $\ell:\N\to\N$ and a universal constant $C$ such that every $k$-planar graph is contained in $H\boxtimes P \boxtimes K_{\ell(k)}$ for some graph $H$ with treewidth at most $C$?  Perhaps $C=3$, which is the case for planar graphs (\cref{PlanarProduct}(b)) and for $1$-planar graphs (\cref{1_planar_product}). Note that $C\geq 3$ even for planar graphs \citep{dujmovic.joret.ea:planar}.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar,UWY22,DHHW22}]\label{GenusProduct}
  Every graph of Euler genus $g$ is contained in:
  \begin{compactenum}[(a)]
  \item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
  \item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some planar graph $H$ of treewidth at most $3$ and for some path $P$.
  \end{compactenum}
\end{thm}

@@

\begin{thm}
\label{gkPlanarProduct}
Every $(g,k)$-planar graph is contained in $H\boxtimes P \boxtimes K_\ell$ for some graph $H$ with $\tw(H) \leq \binom{k+4}{3}-1$, where $\ell:=\max\{2g,3\}\cdot(6k^2+16k+10)$.
\end{thm}

@@

%\note{PM}{To make the referee happy I've added the ``bounded-degree'' qualifier to string graphs and the ``bounded clique number'' qualifier to map graphs.  Do we want to explcitly state that the clique number of a $(g,d)$-map graph is bounded by a function of $g$ and $d$?  Proof (for the $g=0$ case): A $d$-map representation gives a drawing with at most $\floor{\tfrac{d-2}{2}}\ceil{\tfrac{d-2}{2}}\in O(d^2)$ crossings per edge, but any drawing of $K_{\omega}$ on a surface of genus-$g$ has an edge that is crossed $\Omega(\omega^2)$ times. So if $G$ is a $d$-map graph then $\omega(G) \in O(d)$.} \note{DW}{I see no need to state that the clique number of a $(g,d)$-map graph is bounded by a function of $g$ and $d$. This is implied by \cref{gdMapProduct} and is not used in its proof. Also, surely there is a more combinatorial proof (with better bounds) than what you get by counting crossings.}

@@

For any integer $d\ge 3$, an $\R^2$-embedded (multi)graph $G$ is \defin{$d$-framed} if it has a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in some face (possibly the outer face) of $G_0$. The embedded (multi)graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.  A $d$-framed graph $G$ is \defin{edge-maximal} with respect to its frame $G_0$ if, for each face $F$ of $G_0$ and each pair of distinct vertices $v,w\in V(F)$, $vw\in E(F)$ or $G$ contains an edge $vw$ whose interior is contained in $F$.

@@

A \defin{BFS spanning forest} $T$ of $G$ rooted at a set $V_0\subseteq V(G)$ is a spanning forest of $G$ that contains a tree rooted at each vertex of $V_0$ with the property that for each $v\in V(G)$,
\[  \min\{\dist_G(v,w):w\in V_0\}=\min\{\dist_T(v,w):w\in V_0\}.  \]
\cref{d_framed_product_stucture} is a consequence of the following technical lemma, which is an extension of the analogous result for plane graphs~\cite[Lemma~17]{dujmovic.joret.ea:planar}.  For readers already familiar with the proof in \cite{dujmovic.joret.ea:planar}, the main difference here is that a typical tripod in the proof of \cite[Lemma~17]{dujmovic.joret.ea:planar} consists of three upward paths in a BFS tree/forest whose lower endpoints belong to a single triangular face $\tau$.  In the current setting, a typical tripod consists of three upward paths whose lower endpoints belong to a common face $\tau$ of $G_0$ plus all of the (at most $d$) vertices of $\tau$ that are not already included in tripods.

@@

\begin{proof}
This proof is similar to the proofs of Lemmas~13 and 17 in \cite{dujmovic.joret.ea:planar}, but is complicated by several factors.  In particular, the reader should keep in mind that $G$ and $G_0$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

@@

  \begin{figure}[!h]
    \begin{center}
      \begin{tabular}{c@{}c@{}c}
        \includegraphics{figs/zoomba-1} &
        \includegraphics{figs/zoomba-2} &
        \includegraphics{figs/zoomba-3} \\
        \multicolumn{3}{c}{{\color{brew8}\raisebox{-3pt}{\rule{12pt}{12pt}}} $S$\qquad {\color{brew2}\raisebox{-3pt}{\rule{12pt}{12pt}}} $X$} \\
        $k=1$ & $k=2$ & $k=3$
      \end{tabular}
    \end{center}
    \caption{Three cases when choosing the face $\tau$ of $N_0$.}
    \label{boring_figure}
  \end{figure}

@@

\begin{proof}
Let $G$ be a (connected) $1$-plane multigraph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}. We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If distinct vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no newly-introduced edge $vw$ is on the boundary of a face with an edge parallel to $vw$.  Since $G$ is connected, each edge added during this process splits a face $F$ with a facial walk of length at least $4$ into two faces each having a shorter facial walk than that of $F$. Therefore this process eventually reaches a stage where each face is either bounded by a $3$-cycle or a (pre-existing) $2$-cycle, and the process ends.

@@

For an integer $k\geq 0$, a \defin{$k$-queue layout} of a graph $G$ consists of a linear ordering $\preceq$ of $V(G)$ and a partition $\{E_1,E_2,\dots,E_k\}$ of $E(G)$, such that for $i\in\{1,2,\dots,k\}$, no two edges in $E_i$ are nested with respect to $\preceq$. That is, it is not the case that $v\prec x \prec y \prec w$ for edges $vw,xy\in E_i$. The \defin{queue-number} of a graph $G$, denoted by \defin{$\qn(G)$}, is the minimum integer $k$ such that $G$ has a $k$-queue layout. Queue-number was introduced by \citet{HLR92}, who famously conjectured that planar graphs have bounded queue-number. \citet{dujmovic.joret.ea:planar} proved this conjecture using \cref{PlanarProduct} and the following lemma. Indeed, resolving this question was the motivation for the development of \cref{PlanarProduct}.

\begin{lem}[\citep{dujmovic.joret.ea:planar}]
\label{qn}
If $G\subseteq H \boxtimes P \boxtimes K_\ell$ then
$\qn(G) \leq  3 \ell \, \qn(H) + \floor{\tfrac{3}{2}\ell}
\leq 3 \ell \, 2^{\tw(H)}  - \ceil{\tfrac{3}{2}\ell}$.
\end{lem}

@@

Note that \citet{dujmovic.joret.ea:planar} previously proved the bound of
$O(g^{k+2})$ for $(g,k)$-planar graphs using \cref{GenusProduct} and an ad-hoc method. Our result provides a better bound when $g>2^{k^2}$.

@@

For planar graphs of treewidth at most $3$ a $O(p^2\log p)$ upper bound is known:
